### 2023-10-04 doc/_bayesian_inference.qmd
update figure numbering

To overcome this computational limitation, we rely on the recent development of
the automatic differentiation variational inference algorithm (ADVI)
[@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =
(\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
replaced by an approximate posterior distribution $q_\phi(\theta)$, where $\phi$
fully parametrizes the approximate distribution. As further explained in
@sec-vi_primer, the numerical integration problem is replaced by an optimization
problem of the form
$$
q^*_\phi(\theta) = \min _\phi
D_{KL}(
    q_\phi(\theta) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
$${#eq-vi_objective}
where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the
complicated high-dimensional numerical integration problem is transformed into a
much simpler problem of finding the value of the parameters $\phi$ such that
@eq-vi_objective is satisfied as best as possible within some finite computation
time. Although to compute @eq-vi_objective, we require the posterior
distribution we are trying to approximate $\pi(\theta \mid
\underline{\underline{R}})$, it can be shown that maximizing the so-called
evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the
variational free energy [@buckley2017]---is mathematically equivalent to
performing the optimization prescribed by @eq-vi_objective. We direct the reader
to @sec-vi_primer for a short primer on variational inference.

>>>>

To overcome this computational limitation, we rely on the recent development of
the automatic differentiation variational inference algorithm (ADVI)
[@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =
(\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
replaced by an approximate posterior distribution $q_\phi(\theta)$, where $\phi$
fully parametrizes the approximate distribution. As further explained in
@sec-vi_primer, the numerical integration problem is replaced by an optimization
problem of the form
$$
q^*_\phi(\theta) = \min _\phi
D_{KL}(
    q_\phi(\theta) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
$${#eq-vi_objective}
where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the
complicated high-dimensional numerical integration problem is transformed into a
much simpler problem of finding the value of the parameters $\phi$ such that
@eq-vi_objective is satisfied as best as possible within some finite computation
time. Although to compute @eq-vi_objective, we require the posterior
distribution we are trying to approximate $\pi(\theta \mid
\underline{\underline{R}})$, it can be shown that maximizing the so-called
evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the
variational free energy [@gottwald2020]---is mathematically equivalent to
performing the optimization prescribed by @eq-vi_objective. We direct the reader
to @sec-vi_primer for a short primer on variational inference.

### 2023-10-04 doc/_hierarchical_genotypes_results.qmd
update figure numbering

Hierarchical models can also capture another experimental design in which
multiple barcodes map to the same or an equivalent genotype. As we will show,
this many-to-one mapping can improve the inference compared to the extreme cases
of inferring the fitness of each barcode independently or pooling the data of
all barcodes mapping to a single genotype. As schematized in @fig-05(A), a small
modification of the base model allows us to map the structure of our original
model to that of a hierarchical model with a fitness hyperparameter vector
$\underline{\theta}^G$, where $G$ is the number of genotypes in the dataset.

@fig-05(B) shows a single experimental replicate in which 90 genotypes were
assigned a random number of barcodes (a multinomial distribution with a mean of
ten barcodes per genotype) for a total of 900 non-neutral barcodes. To assess
the performance of the hierarchical model proposed in @fig-05(A), we performed
inference using this hierarchical model, as well as the two extreme cases of
ignoring the connection between the barcodes belonging to the same
genotype---equivalent to performing inference using the model presented in
@fig-02(A) over the barcodes---or pooling the data of all barcodes belonging to
the same genotype into a single count---equivalent to performing inference using
the model presented in @fig-02(A) over the pooled barcodes. @fig-05(C-D) shows
the comparison between the simulation ground truth and the inferred values for
these three cases. Not only do the hierarchical model results show higher
degrees of correlation with the ground truth, but the error bars (representing
the 68\% credible regions) are smaller, meaning that the uncertainty in the
estimate of the parameter we care about decreases when using the hierarchical
model. The improvement in the prediction can be seen in @fig-05(F) where the
empirical cumulative distribution function of the absolute difference between
the mean inferred value and the simulation ground truth is shown for all three
inference models. The hierarchical model's curve ascends more rapidly, showing
that, in general, the inferred values are closer to the ground truth. For
completeness, @fig-05(G) shows some examples of how the hierarchical model can
capture the raw log-frequency count observations.

![**Hierarchical model for multiple barcodes per genotype.** (A) Schematic
depiction of the hierarchical structure for multiple barcodes mapping to a
single genotype. A set of barcodes mapping to an equivalent genotype map to
"local" fitness values $s^{(b)}$ that are connected via a hyper-fitness
parameter for the genotype $\theta^{(g)}$. (B) Simulated dataset with 100
neutral lineages and 900 barcodes of interest distributed among 90 genotypes.
(C-E) Comparison between the inferred and ground truth fitness values for a
hierarchical model (C), a model where each barcode is inferred independently
(D), and a model where barcodes mapping to the same genotype are pooled
together (E). Gray error bars represent the 68\% credible regions. (F) Empirical
cumulative distribution function (ECDF) of the absolute difference between the
inferred mean and the ground truth fitness values for all three models. (G)
Examples of the posterior predictive checks for all neutral lineages (upper left
panels) and a subset of representative mutant lineages. Shaded regions from
light to dark represent the 95\%, 68\%, and 5\% credible
regions.](./figs/fig05){#fig-05}

>>>>

Hierarchical models can also capture another experimental design in which
multiple barcodes map to the same or an equivalent genotype. As we will show,
this many-to-one mapping can improve the inference compared to the extreme cases
of inferring the fitness of each barcode independently or pooling the data of
all barcodes mapping to a single genotype. As schematized in @fig-06(A), a small
modification of the base model allows us to map the structure of our original
model to that of a hierarchical model with a fitness hyperparameter vector
$\underline{\theta}^G$, where $G$ is the number of genotypes in the dataset.

@fig-06(B) shows a single experimental replicate in which 90 genotypes were
assigned a random number of barcodes (a multinomial distribution with a mean of
ten barcodes per genotype) for a total of 900 non-neutral barcodes. To assess
the performance of the hierarchical model proposed in @fig-06(A), we performed
inference using this hierarchical model, as well as the two extreme cases of
ignoring the connection between the barcodes belonging to the same
genotype---equivalent to performing inference using the model presented in
@fig-02(A) over the barcodes---or pooling the data of all barcodes belonging to
the same genotype into a single count---equivalent to performing inference using
the model presented in @fig-02(A) over the pooled barcodes. @fig-06(C-D) shows
the comparison between the simulation ground truth and the inferred values for
these three cases. Not only do the hierarchical model results show higher
degrees of correlation with the ground truth, but the error bars (representing
the 68\% credible regions) are smaller, meaning that the uncertainty in the
estimate of the parameter we care about decreases when using the hierarchical
model. The improvement in the prediction can be seen in @fig-06(F) where the
empirical cumulative distribution function of the absolute difference between
the mean inferred value and the simulation ground truth is shown for all three
inference models. The hierarchical model's curve ascends more rapidly, showing
that, in general, the inferred values are closer to the ground truth. For
completeness, @fig-06(G) shows some examples of how the hierarchical model can
capture the raw log-frequency count observations.

![**Hierarchical model for multiple barcodes per genotype.** (A) Schematic
depiction of the hierarchical structure for multiple barcodes mapping to a
single genotype. A set of barcodes mapping to an equivalent genotype map to
"local" fitness values $s^{(b)}$ that are connected via a hyper-fitness
parameter for the genotype $\theta^{(g)}$. (B) Simulated dataset with 100
neutral lineages and 900 barcodes of interest distributed among 90 genotypes.
(C-E) Comparison between the inferred and ground truth fitness values for a
hierarchical model (C), a model where each barcode is inferred independently
(D), and a model where barcodes mapping to the same genotype are pooled
together (E). Gray error bars represent the 68\% credible regions. (F) Empirical
cumulative distribution function (ECDF) of the absolute difference between the
inferred mean and the ground truth fitness values for all three models. (G)
Examples of the posterior predictive checks for all neutral lineages (upper left
panels) and a subset of representative mutant lineages. Shaded regions from
light to dark represent the 95\%, 68\%, and 5\% credible
regions.](./figs/fig06){#fig-06}

### 2023-10-04 doc/_hierarchical_replicates_results.qmd
update figure numbering

Hierarchical models present a middle ground between these extremes. First,
hierarchical models rely on the definition of so-called *hyper-parameters*, that
capture the parametric inference we are interested in---for this inference
problem, we have a hyper-fitness value $\theta^{(m)}$ for each non-neutral
barcode. Second, each experiment draws randomly from the distribution of this
hyper-parameter, allowing for subtle variability between experiments to be
accounted for---in the present inference pipeline, each experimental replicate
gets assigned a *local* fitness value $s^{(m)}_i$, where the extra sub-index
indicates the $i$-th experimental replicate. For our hierarchical model, these
local fitness values interact via the common dependency on the global
hyper-fitness value. As schematized in @fig-03(A), this interaction runs both
ways. First, the data from one experimental replicate
($\underline{\underline{R}}^M_k$ in the diagram) informs all local fitness
values via the global hyper-fitness (upper panel in @fig-03(A)). Second, the
local fitness value is informed by the data from all experimental replicates via
the same global hyper-fitness parameter (lower panel in @fig-03(A)).

To test the performance of this model, we simulated two experimental replicates
with 1000 unique barcodes (see @fig-03(B-C)) where we randomly sampled a ground
truth hyper-fitness value $\theta^{(m)}$ for each barcode. We sampled a
variation from this hyper-fitness value for each experimental replicate
$s^{(m)}_i$ to capture experimental batch effects. @fig-03(D) shows the
relationship between hyper-fitness and replicate fitness values for this
simulation. The spread around the identity line represents the expected
batch-to-batch variation. The posterior predictive checks examples in @fig-03(E)
show that the hierarchical model can correctly fit the data for each
experimental replicate. Furthermore, @fig-03(F-G) show a high correlation
between the ground truth and the inferred fitness values. The empirical
cumulative distribution functions shown in @fig-03(H-I) reveal that for $\approx
75\%$ of the non-neutral barcodes, the ground truth hyper-fitness values fall
within one standard deviation from the mean value in the posterior
distributions.

![**Hierarchical model on experimental replicates**. (A) Schematic depiction of
the interactions between local fitness values $\underline{s}_k^M$ through the
global hyper-fitness value $\underline{\theta}^M$ for $K$ hypothetical
experimental replicates. The upper diagram shows how the data from replicate $k$
informs all local fitness values via the hyper-fitness parameter. The lower
panel shows the reverse, where all other datasets inform the local fitness
value. (B-C) Simulated replicate datasets with 900 barcodes of interest and 100
neutral lineages. (D) Comparison between the simulation ground truth
hyper-fitness and each replicate ground truth fitness. The scatter between
parameters captures experimental batch effects. (E) Examples of the posterior
predictive checks for all neutral lineages (upper left panels) and a subset of
representative mutant lineages. Shaded regions from light to dark represent the
95\%, 68\%, and 5\% credible regions. (F-G) Comparison between the simulation's
ground truth hyper-fitness (F) and replicate fitness (G) values and the inferred
parameters. Gray error bars represent the 68\% posterior credible region. (H-I)
The empirical cumulative distribution function (ECDF) for the absolute z-score
value of the ground truth parameter value within the inferred hyper-fitness
posterior distribution (H) and replicate fitness (I).](./figs/fig03){#fig-03}

As shown in @fig-04, the structure imposed by the hierarchical model schematized
in @fig-03(A), where we explicitly account for the connection between
experimental replicates can improve the quality of the inference. Inferred
fitness values between experimental replicates exhibit a stronger degree of
correlation in the hierarchical model (@fig-04(A)) compared to conducting
inference on each replicate independently (@fig-04(B)). Moreover, when comparing
the inferred hyper-fitness values---the objective parameter when performing
multiple experimental measurements---the hierarchical model outperforms
averaging the independent experimental replicates as shown in @fig-04(C) and
(D).

![**Comparison between hierarchical model and single dataset model**. (A-B)
comparison of inferred fitness values between experimental replicates when
fitting a hierarchical model (A) or independently fitting each dataset (B). Gray
error bars represent the 68\% credible regions. (C) Comparison between the
ground truth hyper-fitness value and the inferred parameters. The blue dots show
the inferred hyper-fitness values when assuming a hierarchical model. Gray error
bars show the 68\% credible region for this inference. The yellow dots show the
average of the mean inferred fitness values for the two experimental replicates.
No error bars are shown for these, as it is inappropriate to compute one with
two data points per non-neutral barcode. (D) Empirical cumulative distribution
function (ECDF) of the absolute difference between the inferred mean and the
ground truth hyper-fitness.](./figs/fig04){#fig-04}

>>>>

Hierarchical models present a middle ground between these extremes. First,
hierarchical models rely on the definition of so-called *hyper-parameters*, that
capture the parametric inference we are interested in---for this inference
problem, we have a hyper-fitness value $\theta^{(m)}$ for each non-neutral
barcode. Second, each experiment draws randomly from the distribution of this
hyper-parameter, allowing for subtle variability between experiments to be
accounted for---in the present inference pipeline, each experimental replicate
gets assigned a *local* fitness value $s^{(m)}_i$, where the extra sub-index
indicates the $i$-th experimental replicate. For our hierarchical model, these
local fitness values interact via the common dependency on the global
hyper-fitness value. As schematized in @fig-04(A), this interaction runs both
ways. First, the data from one experimental replicate
($\underline{\underline{R}}^M_k$ in the diagram) informs all local fitness
values via the global hyper-fitness (upper panel in @fig-04(A)). Second, the
local fitness value is informed by the data from all experimental replicates via
the same global hyper-fitness parameter (lower panel in @fig-04(A)).

To test the performance of this model, we simulated two experimental replicates
with 1000 unique barcodes (see @fig-04(B-C)) where we randomly sampled a ground
truth hyper-fitness value $\theta^{(m)}$ for each barcode. We sampled a
variation from this hyper-fitness value for each experimental replicate
$s^{(m)}_i$ to capture experimental batch effects. @fig-04(D) shows the
relationship between hyper-fitness and replicate fitness values for this
simulation. The spread around the identity line represents the expected
batch-to-batch variation. The posterior predictive checks examples in @fig-04(E)
show that the hierarchical model can correctly fit the data for each
experimental replicate. Furthermore, @fig-04(F-G) show a high correlation
between the ground truth and the inferred fitness values. The empirical
cumulative distribution functions shown in @fig-04(H-I) reveal that for $\approx
75\%$ of the non-neutral barcodes, the ground truth hyper-fitness values fall
within one standard deviation from the mean value in the posterior
distributions.

![**Hierarchical model on experimental replicates**. (A) Schematic depiction of
the interactions between local fitness values $\underline{s}_k^M$ through the
global hyper-fitness value $\underline{\theta}^M$ for $K$ hypothetical
experimental replicates. The upper diagram shows how the data from replicate $k$
informs all local fitness values via the hyper-fitness parameter. The lower
panel shows the reverse, where all other datasets inform the local fitness
value. (B-C) Simulated replicate datasets with 900 barcodes of interest and 100
neutral lineages. (D) Comparison between the simulation ground truth
hyper-fitness and each replicate ground truth fitness. The scatter between
parameters captures experimental batch effects. (E) Examples of the posterior
predictive checks for all neutral lineages (upper left panels) and a subset of
representative mutant lineages. Shaded regions from light to dark represent the
95\%, 68\%, and 5\% credible regions. (F-G) Comparison between the simulation's
ground truth hyper-fitness (F) and replicate fitness (G) values and the inferred
parameters. Gray error bars represent the 68\% posterior credible region. (H-I)
The empirical cumulative distribution function (ECDF) for the absolute z-score
value of the ground truth parameter value within the inferred hyper-fitness
posterior distribution (H) and replicate fitness (I).](./figs/fig04){#fig-04}

As shown in @fig-05, the structure imposed by the hierarchical model schematized
in @fig-04(A), where we explicitly account for the connection between
experimental replicates can improve the quality of the inference. Inferred
fitness values between experimental replicates exhibit a stronger degree of
correlation in the hierarchical model (@fig-05(A)) compared to conducting
inference on each replicate independently (@fig-05(B)). Moreover, when comparing
the inferred hyper-fitness values---the objective parameter when performing
multiple experimental measurements---the hierarchical model outperforms
averaging the independent experimental replicates as shown in @fig-05(C) and
(D).

![**Comparison between hierarchical model and single dataset model**. (A-B)
comparison of inferred fitness values between experimental replicates when
fitting a hierarchical model (A) or independently fitting each dataset (B). Gray
error bars represent the 68\% credible regions. (C) Comparison between the
ground truth hyper-fitness value and the inferred parameters. The blue dots show
the inferred hyper-fitness values when assuming a hierarchical model. Gray error
bars show the 68\% credible region for this inference. The yellow dots show the
average of the mean inferred fitness values for the two experimental replicates.
No error bars are shown for these, as it is inappropriate to compute one with
two data points per non-neutral barcode. (D) Empirical cumulative distribution
function (ECDF) of the absolute difference between the inferred mean and the
ground truth hyper-fitness.](./figs/fig05){#fig-05}

### 2023-10-04 doc/_multienv_results.qmd
update figure numbering

One of the benefits of Bayesian methods is the intrinsic error propagation
embedded in the mathematical framework. For example, the uncertainty on the
value of the population mean fitness values $\underline{\bar{s}}_T$ is
propagated to the relative fitness of a particular strain $s^{(b)}$ since we
defined a joint posterior distribution over all parameters as fully expressed in
@eq-bayes_full. This natural error propagation can help us with the experimental
design schematized in @fig-06(A). Here, rather than performing growth-dilution
cycles in the same environment, the cells are diluted into a different 
environment. Thus, the uncertainty on the fitness estimate for the previous
environment must be propagated to that of the next one.

To validate the extension of our statistical model to this scenario, @fig-06(B)
shows the trajectory of the log frequency ratios between adjacent time points.
The different colored regions correspond to the different environments. For this
simulation, the growth rate of Environment 2 was set to be, on average, half of
the average growth rate in Environment 1. Equivalently, the growth rate in
Environment 3 was set to be, on average, twice the average growth rate in
Environment 1. @fig-06(C-E) show the correspondence between the simulation
ground truth and the inferred fitness values, where the error bars represent the
68\% credible region. @fig-06(F) summarizes the performance of our inference
pipeline by showing the empirical cumulative distribution functions for the
absolute value of the ground truth fitness value z-score within the posterior
distribution. This plot shows that, overall, $\approx 75\%$ of inferred mean
values fall within one standard deviation of the ground truth. For completeness,
@fig-06(G) shows the posterior predictive checks for a few example barcodes.

![**Multi-environment fitness inference.** (A) Schematic of the simulated
experimental design where growth-dilution cycles are performed into different
environments for each cycle. (B) log frequency rations between adjacent time
points. Darker lines represent the neutral barcodes. The colors in the
background demark the corresponding environment, matching colors in (A).
Environment 2 is set to have, on average, half the growth rate of environment 1.
Likewise, environment 3 is set to have, on average, twice the growth rate of
environment 1. (C-E) Comparison between the ground truth fitness value from the
logistic-growth simulation and the inferred fitness value for each environment.
Gray error bars represent the 68\% posterior credible region. (F) The empirical
cumulative distribution function (ECDF) for the absolute z-score value of the
ground truth parameter value within the inferred fitness posterior distribution
for all fitness values (black line) and each environment individually (color
lines). (G) Examples of the posterior predictive checks for all neutral lineages
(upper left panel) and a subset of representative mutant lineages. Shaded
regions surrounding the data represent the 95\%, 68\%, and 5\% credible regions
for the data. The reported errors above the plot represent the 68\% credible
region on the mutant relative fitness marginal distribution. Background colors
match those of (A).](./figs/fig06){#fig-06}

>>>>

One of the benefits of Bayesian methods is the intrinsic error propagation
embedded in the mathematical framework. For example, the uncertainty on the
value of the population mean fitness values $\underline{\bar{s}}_T$ is
propagated to the relative fitness of a particular strain $s^{(b)}$ since we
defined a joint posterior distribution over all parameters as fully expressed in
@eq-bayes_full. This natural error propagation can help us with the experimental
design schematized in @fig-03(A). Here, rather than performing growth-dilution
cycles in the same environment, the cells are diluted into a different 
environment. Thus, the uncertainty on the fitness estimate for the previous
environment must be propagated to that of the next one.

To validate the extension of our statistical model to this scenario, @fig-03(B)
shows the trajectory of the log frequency ratios between adjacent time points.
The different colored regions correspond to the different environments. For this
simulation, the growth rate of Environment 2 was set to be, on average, half of
the average growth rate in Environment 1. Equivalently, the growth rate in
Environment 3 was set to be, on average, twice the average growth rate in
Environment 1. @fig-03(C-E) show the correspondence between the simulation
ground truth and the inferred fitness values, where the error bars represent the
68\% credible region. @fig-03(F) summarizes the performance of our inference
pipeline by showing the empirical cumulative distribution functions for the
absolute value of the ground truth fitness value z-score within the posterior
distribution. This plot shows that, overall, $\approx 75\%$ of inferred mean
values fall within one standard deviation of the ground truth. For completeness,
@fig-03(G) shows the posterior predictive checks for a few example barcodes.

![**Multi-environment fitness inference.** (A) Schematic of the simulated
experimental design where growth-dilution cycles are performed into different
environments for each cycle. (B) log frequency rations between adjacent time
points. Darker lines represent the neutral barcodes. The colors in the
background demark the corresponding environment, matching colors in (A).
Environment 2 is set to have, on average, half the growth rate of environment 1.
Likewise, environment 3 is set to have, on average, twice the growth rate of
environment 1. (C-E) Comparison between the ground truth fitness value from the
logistic-growth simulation and the inferred fitness value for each environment.
Gray error bars represent the 68\% posterior credible region. (F) The empirical
cumulative distribution function (ECDF) for the absolute z-score value of the
ground truth parameter value within the inferred fitness posterior distribution
for all fitness values (black line) and each environment individually (color
lines). (G) Examples of the posterior predictive checks for all neutral lineages
(upper left panel) and a subset of representative mutant lineages. Shaded
regions surrounding the data represent the 95\%, 68\%, and 5\% credible regions
for the data. The reported errors above the plot represent the 68\% credible
region on the mutant relative fitness marginal distribution. Background colors
match those of (A).](./figs/fig03){#fig-03}
